---
title: "Fusil Analysis Exploring the Viability and Paralogy Datasets"
author: "Hussein"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```


```{r echo=TRUE, message=FALSE, warning=FALSE}

library(tidyverse)
library(readxl)
library(writexl)

```


## Loading the Necessary Data sets



```{r echo=TRUE, message=FALSE, warning=FALSE}
#Get a list of the protein coding genes 

protein_coding_genes <- read_delim("C:/Users/HP-ssd/Desktop/Short term project/protein coding genes/gene_with_protein_product.txt", 
                                   delim = "\t", escape_double = FALSE, 
                                   trim_ws = TRUE)

protein_coding_genes_list <- protein_coding_genes$symbol


#Get FUSIL file

fusil_m_gene <-  read_delim("C:/Users/HP-ssd/Desktop/Short term project2/fusil.csv")

length(unique(fusil_m_gene$gene_symbol))

#Get a list of Gene Paralogues from biomart

human_gene_paralogues <- read.csv("C:/Users/HP-ssd/Desktop/Short term project2/paralogues/human_gene_paralogues.csv")

human_gene_paralogues <- human_gene_paralogues %>%
  select(-1,-2,-4)%>%
  rename(gene_symbol = external_gene_name )

length(unique(human_gene_paralogues$gene_symbol))

#Merge the FUSIL file with paralogues dataset

paralogue_fusil <- human_gene_paralogues %>%
  left_join(fusil_m_gene)%>%
  select(-6, -7) %>%
  mutate(hsapiens_paralog_associated_gene_name = na_if(hsapiens_paralog_associated_gene_name,"")) %>%
  distinct()

length(unique(paralogue_fusil$gene_symbol))

```


## Descriptive Analysis


### Claculating Percentage of Genes in Each FUSIL bin


```{r echo=TRUE, message=FALSE, warning=FALSE}

# Gene Count (Percentage) distribution across FUSIL bins -----------------------------


gene_counts <- fusil_m_gene %>%
  count(fusil) %>%
  mutate(percentage = (n/sum(n)*100)) %>%
  mutate(total_gene_count =n)


gene_counts$fusil <- factor(gene_counts$fusil, 
                                         levels = c("CL", "DL", "SV", "VP", "VnP"  ))

ggplot(gene_counts, aes(x= fusil, y=percentage, fill = fusil))+
  geom_bar(stat = "identity", position = "dodge")+
  labs(y = "Percentage of Genes", x = "FUSIL bin", title = "Percentages of Genes by FUSIL Category") +
  theme_minimal()+
  scale_fill_manual(
    values = c(
      "CL" = "#E41A1C",
      "DL" = "#377EB8",
      "SV" = "#4DAF4A",
      "VnP" = "#984EA3",
      "VP" = "#FF7F00"
    ))
```



### Comparing Gene counts per FUSIL between those with and without paralogues


```{r echo=TRUE, message=FALSE, warning=FALSE}

#Count of genes with paralogues and without paralogues per FUSIL bin


na_paralogue_count <- paralogue_fusil %>%
  group_by(gene_symbol) %>%
  filter(all(is.na(hsapiens_paralog_perc_id))) %>%
  ungroup() %>%
  distinct() %>%
  count(fusil) %>%
  mutate(gene_with_no_paralogue_count =n)

sum(na_paralogue_count$n)

gene_with_paralogue_count <- paralogue_fusil %>%
  group_by(gene_symbol) %>%
  filter(!all(is.na(hsapiens_paralog_perc_id))) %>%
  ungroup() %>%
  distinct() %>%
  na.omit() %>%
  select(1,6) %>%
  distinct()%>%
  count(fusil) %>%
  mutate(gene_with_paralogue_count =n)
  

sum(gene_with_paralogue_count$n)


Summary_gene_count <- gene_counts %>%
  left_join(na_paralogue_count, by = "fusil") %>%
  left_join(gene_with_paralogue_count, by = "fusil") %>%
  select(1,4,6,8)



Summary_gene_count_long <- Summary_gene_count %>%
  pivot_longer(cols = c(gene_with_paralogue_count, gene_with_no_paralogue_count),
               names_to = "ParalogueStatus",
               values_to = "Count")



Summary_gene_count_long$fusil <- factor(Summary_gene_count_long$fusil, 
                            levels = c("CL", "DL", "SV", "VP", "VnP" ))

df_totals <- Summary_gene_count %>%
  select(fusil, total_gene_count) %>%
  distinct()

ggplot(Summary_gene_count_long, aes(x= fusil, y= Count, fill = ParalogueStatus))+
  geom_bar(stat = "identity") +
  geom_text(data = df_totals, aes(x = fusil, y = total_gene_count, label = total_gene_count),
            vjust = -0.5, inherit.aes = FALSE) +
  labs(
    title = "Gene Counts by Paralogue Presence",
    x = "FUSIL bin", y = "Gene Count"
  )



```



### Calculatign the Average Number of Paralogues per Gene for each FUISL BIN Across Similarity Thresholds


```{r echo=TRUE, message=FALSE, warning=FALSE}

paralogue_fusil_filtered <- paralogue_fusil %>%
  na.omit() %>% # remove null values
  filter(hsapiens_paralog_associated_gene_name %in% protein_coding_genes_list) # filter for only protein coding


length(unique(paralogue_fusil_filtered$gene_symbol))



paralogue_fusil_filtered$fusil <- factor(paralogue_fusil_filtered$fusil, 
                                         levels = c("VnP", "VP", "SV", "DL","CL"  )) # keep categories in order

thresholds <- c(30,50,70)     # Define thresholds


fusil_cat_thresh <-  lapply(thresholds, function(thresh) {
  paralogue_fusil_filtered %>%
    filter(hsapiens_paralog_perc_id >= thresh) %>%
    group_by(gene_symbol, fusil) %>%
    tally(name = "Paralogues_above_threshold") %>%
    mutate(Threshold =thresh)
}) %>%
  bind_rows()

length(unique(fusil_cat_thresh$gene_symbol))


summary_thresh <- fusil_cat_thresh %>%
  group_by(fusil, Threshold) %>%
  summarise(avg_paralogues = mean(Paralogues_above_threshold), .groups= "drop")

summary_thresh$fusil <- factor(summary_thresh$fusil, 
                            levels = c("CL", "DL", "SV", "VP", "VnP"  ))

ggplot(summary_thresh, aes(x= factor(Threshold), y=avg_paralogues, fill = fusil))+
  geom_bar(stat = "identity", position = "dodge")+
  facet_wrap(~Threshold)+
  labs(x = "Homology Threshold (%)", y = "Average Number of Paralogues", title = "Paralogue Count by FUSIL Category") +
  theme_minimal()+
  theme(axis.text.x = element_blank())+
  scale_fill_manual(
    values = c(
      "CL" = "#E41A1C",
      "DL" = "#377EB8",
      "SV" = "#4DAF4A",
      "VnP" = "#984EA3",
      "VP" = "#FF7F00"
    ))

### Plotting to see the paralogues counts across % similarity

paralogue_fusil_filtered$fusil <- factor(paralogue_fusil_filtered$fusil, 
                                         levels = c("CL", "DL", "SV", "VP", "VnP" ))
                                         

ggplot(paralogue_fusil_filtered, aes( x= hsapiens_paralog_perc_id, fill = fusil ))+
  geom_density(alpha = 0.5) +
  labs(title = "Distribution of Paralogue across Similarity thresholds by FUSIL Category",
       x = "Similarity Score",
       y = "Density")+
  theme_minimal()


```



### Generating a binary matrix of paralogue presence while accounting for protein-coding genes



```{r echo=TRUE, message=FALSE, warning=FALSE}

# # Binary data for paralogue presence among FUSIL Categories --------


library(stringr)

has_paralogues_above_thresholds <- function(gene, paralogue_perc_id, pc_genes, threshold) {
  
  if (!(gene %in% pc_genes)) {
    return(0)
  }
  
  if(is.na(paralogue_perc_id) || paralogue_perc_id == "") {
    return(0)
  }
  
  perc <- as.numeric(str_extract(paralogue_perc_id, "\\d+\\.?\\d*")[[1]])
  
  if(!is.na(perc) && perc >= threshold ) {
    return(1)
  } else {
    return(0)
  }
}


thresholds <- c(30, 50, 70)

pc_genes <- protein_coding_genes$symbol


binary_thresh <- lapply(thresholds, function (thresh) {
 
  binary_column <- mapply(has_paralogues_above_thresholds,
                          gene = paralogue_fusil$gene_symbol,
                         paralogue_perc_id = paralogue_fusil$hsapiens_paralog_perc_id,
                         MoreArgs = list(pc_genes = pc_genes, threshold = thresh)
                         )
  col_name <- paste0("Has_paralogue_over_", thresh)
  setNames(data.frame(binary_column), col_name)
}
  
  )

binary_columns_df <- bind_cols(binary_thresh)

fusil_cat_thresh_2 <- bind_cols(paralogue_fusil, binary_columns_df)

length(unique(fusil_cat_thresh_2$gene_symbol))

#write.csv(fusil_cat_thresh_2, "C:/Users/HP-ssd/Desktop/fusil_cat_thresh_3.csv")

binary_summary_per_gene <- fusil_cat_thresh_2 %>%
  select(gene_symbol, fusil, starts_with("Has_paralogue_over_")) %>%
  group_by(gene_symbol, fusil) %>%
  summarise(across(starts_with("Has_paralogue_over_"), max, na.rm = TRUE), .groups = "drop")


#write.csv(binary_summary_per_gene, "C:/Users/HP-ssd/Desktop/binary_summary_per_gene.csv")

count_summary_per_gene <- fusil_cat_thresh_2 %>%
  select(gene_symbol,fusil, starts_with("Has_paralogue_over_")) %>%
  group_by(gene_symbol, fusil) %>%
  summarise(across(starts_with("Has_paralogue_over_"), sum, na.rm = TRUE), .groups = "drop") %>%
  rename_with(~ gsub("Has", "Count", .x), starts_with("Has"))




#Visualise the data

binary_summary_per_gene_long <- binary_summary_per_gene %>%
  pivot_longer( cols = starts_with("Has_paralogue_over_"),
                names_to = "Threshold",
                values_to = "Has_Paralogue"
)


binary_summary_per_gene_long <- binary_summary_per_gene_long %>%
  mutate(Threshold = as.numeric(gsub("Has_paralogue_over_", "", Threshold)))


plot_binary_summary_per_gene_long <- binary_summary_per_gene_long %>%
  group_by(fusil, Threshold, Has_Paralogue ) %>%
  summarise(Count=n(), , .groups = "drop") %>%
  group_by(fusil, Threshold) %>%
  mutate(Proportions = Count/sum(Count))

ggplot(plot_binary_summary_per_gene_long, aes( x = factor(Threshold), y = Proportions, fill = factor(Has_Paralogue)))+
  geom_col(position = "dodge") +
  facet_wrap(~ fusil)+
  labs(
    title = "Proportion of Genes With and Without Paralogue by FUSIL Class",
    x = "Similarity Threshold (%)",
    y = "Proportion of Genes",
    fill = "Has Paralogue"
  ) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("0" = "#FF9999", "1" = "#66CC99"),
                    labels = c("No", "Yes")) +
  theme_minimal(base_size = 14)



ggplot(plot_binary_summary_per_gene_long, aes( x = factor(fusil), y = Proportions, fill = factor(Has_Paralogue)))+
  geom_col(position = "dodge") +
  facet_wrap(~ Threshold)+
  labs(
    title = "Proportion of Genes With and Without Paralogue by FUSIL Class",
    x = "FUSIL bin",
    y = "Proportion of Genes",
    fill = "Has Paralogue"
  ) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("0" = "#FF9945", "1" = "#690C99"),
                    labels = c("No", "Yes")) +
  theme_minimal(base_size = 14)


```



### Comparing the FUSIL bin of Query Genes and their Paralogues
#### Are they in the same category?


```{r echo=TRUE, message=FALSE, warning=FALSE}

#Create a matrix of fusil categories for genes and paralogues

Fusil_genes <- fusil_m_gene %>%
  select(-1,-2)

Fusil_genes_paralogues <- human_gene_paralogues %>%
  left_join( Fusil_genes, by = c( "hsapiens_paralog_associated_gene_name" = "gene_symbol"))%>%
  filter(hsapiens_paralog_associated_gene_name %in% protein_coding_genes_list) %>%
  rename("fusil_paralogue" = "fusil")

Fusil_genes_paralogues <- Fusil_genes_paralogues %>%
  left_join(Fusil_genes, by = c( "gene_symbol" = "gene_symbol"))

Fusil_genes_paralogues <- Fusil_genes_paralogues%>%
  relocate(fusil, .after = "gene_symbol") %>%
  relocate(fusil_paralogue, .after = "hsapiens_paralog_associated_gene_name") %>%
  na.omit()



#ploting

library(ggalluvial) # For Sankey-like alluvial plots
library(reshape2) # for heatmap

df_fusil <- Fusil_genes_paralogues %>%
  count(fusil)

sum(df_fusil$n)

fusil_table <- Fusil_genes_paralogues %>%
  count(fusil, fusil_paralogue) %>%
  left_join(df_fusil, by = c( "fusil" = "fusil")) %>%
  mutate(percentage = n.x*100/ n.y)

sum(fusil_table$n)

ggplot(fusil_table, aes(axis1 = fusil, axis2 = fusil_paralogue, y = percentage))+
  geom_alluvium(aes(fill = fusil),width = 0.2, show.legend = FALSE)+
  geom_stratum(width = 0.2)+
  geom_text(stat = "stratum", aes(label = after_stat(stratum)))+
  scale_x_discrete(limits = c("Gene FUSIL", "Paralog FUSIL"), expand = c(.01, .01)) +
  scale_fill_manual(
    values = c(
      "CL" = "#E41A1C",
      "DL" = "#377EB8",
      "SV" = "#4DAF4A",
      "VnP" = "#984EA3",
      "VP" = "#FF7F00"
    ))+
  theme_minimal() +
  theme(axis.title.y = element_blank(), axis.text.y = element_blank())+
  ggtitle("FUSIL Category Flow: Gene → Paralog")+
  geom_text(aes(label = paste0(round(percentage, 1), "%")),
            stat = "alluvium",
            nudge_x = 0.2,
            size = 3,
            color = "black")
```


### Integrating MCRA Data


```{r echo=TRUE, message=FALSE, warning=FALSE}


# FUSIL gene with MCRA --------------------------------

fusil_m_gene_list <- unique(fusil_m_gene$gene_symbol)

library(biomaRt)

human = useMart("ensembl", dataset = "hsapiens_gene_ensembl")

attributes_human <- listAttributes(human)


fusil_gene_mcra <- getBM(attributes = c("external_gene_name",
                                        "hsapiens_paralog_subtype"),
                               filters = "external_gene_name",
                               values = fusil_m_gene_list,
                               mart = human)


fusil_gene_mcra_df <- fusil_gene_mcra %>%
  filter(!(hsapiens_paralog_subtype == ""))

fusil_gene_mcra_df2 <- fusil_m_gene %>%
  left_join(fusil_gene_mcra_df, by = c ("gene_symbol" ="external_gene_name" )) %>%
  dplyr::select(-1,-2) %>%
  na.omit()


count_mcra_fusil <- fusil_gene_mcra_df2 %>%
  count(hsapiens_paralog_subtype)

broad_subtypes <- c("Bilateria", "Chordata", "Gnathostomata",
                    "Opisthokonta", "Vertebrata" )


mcra_df_fusil <- fusil_gene_mcra_df2  %>%
  filter(hsapiens_paralog_subtype %in% broad_subtypes) %>%
  count(fusil)


mcra_fusil <- fusil_gene_mcra_df2 %>%
  filter(hsapiens_paralog_subtype %in% broad_subtypes) %>%
  group_by(fusil, hsapiens_paralog_subtype) %>%
  tally() %>%
  left_join(mcra_df_fusil, by =c ("fusil" ="fusil")) %>%
  mutate(percentage = (n.x/n.y)*100)


mcra_fusil$hsapiens_paralog_subtype <-factor(mcra_fusil$hsapiens_paralog_subtype, 
                                             levels = c ("Opisthokonta" ,"Bilateria" ,"Chordata","Vertebrata", "Gnathostomata"))

mcra_fusil$fusil <- factor(mcra_fusil$fusil,
                           levels = c("CL", "DL", "SV", "VP", "VnP" ))



ggplot(mcra_fusil, aes( x = factor (hsapiens_paralog_subtype), y = percentage, fill = fusil))+
  geom_col(position = "dodge") +
  labs(
    title = "Gene MCRA subtype distribution by FUSIL Class",
    y = "Percentage of Genes",
    x = "MCRA Subtype Group",
    fill = "Fusil"
  ) +
  theme_minimal(base_size = 14) +
  scale_fill_manual(
    values = c(
      "CL" = "#E41A1C",
      "DL" = "#377EB8",
      "SV" = "#4DAF4A",
      "VnP" = "#984EA3",
      "VP" = "#FF7F00"
    ))



ggplot(mcra_fusil, aes( x = fusil, y = percentage, fill = hsapiens_paralog_subtype))+
  geom_col(position = "dodge") +
  labs(
    title = "FUSIL Gene distribution by MCRA subtype",
    y = "Percentage of Genes",
    x = "FUSIL bin",
    fill = "MCRA Subtype"
  ) +
  theme_minimal(base_size = 14) 

```


